---
title: "Installing SEraster"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing SEraster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6,
  fig.height=4,
  dpi = 72,
  dev = "png"
)
```

To install `SEraster`, we currently recommend using `remotes`:

``` r
require(remotes)
remotes::install_github('JEFworks-Lab/SEraster')
```

We are currently submitting `SEraster` to Bioconductor (this section will be updated after `SEraster` is approved to Bioconductor).

To install this package using Bioconductor, start R (version "4.4.0") and enter:

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SEraster")
```

In addition, `SEraster` is also compatible with `SeuratObject` through `SeuratWrappers`. `SeuratWrappers` implementation can be installed using `remotes`:

``` r
require(remotes)
remotes::install_github('satijalab/seurat-wrappers@SEraster')
```

Documentation and tutorial for the `SeuratWrappers` implementation can be found in the `SEraster` branch of the [`SeuratWrappers` GitHub repository](https://github.com/satijalab/seurat-wrappers/tree/SEraster).
