---
title: "SEraster for Spatial Variable Genes Analysis"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{SEraster for Spatial Variable Genes Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, options, include = FALSE}
library(knitr)
opts_chunk$set(
    cache = FALSE, 
    results = "hold"
)
```

## Introduction
In this tutorial, we will demonstrate how to use `SEraster` as a preprocessing step prior to spatial variable gene analysis with `MERINGUE`.

## Installation

To install `SEraster`, we currently recommend using `remotes`:

```{r}
require(remotes)
remotes::install_github('JEFworks-Lab/SEraster')
```

To install this package using Bioconductor, start R (version "4.4.0") and enter:
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SEraster")
```

## Load libraries
```{r}
suppressMessages(library(SpatialExperiment))
suppressMessages(library(SEraster))
suppressMessages(library(Matrix))
suppressMessages(library(ggplot2))
```

## Load dataset

We will use the MERFISH mPOA (bregma -0.29 slice from a female naive animal) available in the SEraster package. The dataset is formatted as a Spatial Experiment object as shown here: (<https://jef.works/SEraster/reference/merfish_mousePOA.html>).

```{r}
data("merfish_mousePOA")
class(merfish_mousePOA)
```

Let's first plot the total gene expression at the single-cell resolution.

```{r}
## pos is a matrix specifying a cell's x and y positions
pos <- spatialCoords(merfish_mousePOA)
## counts is a genes-by-cells matrix where the gene expression was normalized by cell volume and scaled by 1000
counts <- assays(merfish_mousePOA)$volnorm
df <- data.frame(pos, total_gexp = colSums(counts))
## plot
ggplot(df, aes(x = x, y = y, color = total_gexp)) +
  coord_fixed() +
  scale_color_viridis_c() +
  geom_point(size = 0.5) +
  theme_bw()
  
```

## Rasterize gene expression using SEraster

The `rasterizeGeneExpression` function allows you to rasterize gene expression at a resolution of your choosing. Below is an example of rasterizing gene expression by calculating the mean expression for all cells within each hexagonal pixel at 100 µm resolution.

```{r}
## let's rasterize the dataset at 100 um resolution.
res <- 100
rastGexp <- SEraster::rasterizeGeneExpression(merfish_mousePOA, 
                                              assay_name = "volnorm", 
                                              resolution = res, 
                                              square = FALSE, 
                                              fun = "mean")
SEraster::plotRaster(rastGexp, name = "total_gexp")
```





`SEraster` aggregated 6509 single-cells into 384 pixels, with each pixel containing the normalized gene expression values using mean. You can see how the dimension of `SpatialExperiment` object is changed after rasterization.

```{r}
dim(merfish_mousePOA)
dim(rastGexp)
```

## Perform SVG analysis with `MERINGUE`

`MERINGUE` is a computational framework based on spatial auto-correlation and cross-correlation analysis. It can be used to identify genes with spatially heterogeneous expression patterns. Link to install MERINGUE: (<https://jef.works/MERINGUE/>) First, let's create an adjacency weight matrix for the rasterized dataset.

```{r}
suppressMessages(library(MERINGUE))
## Get neighbor-relationships
rastPos <- spatialCoords(rastGexp)
w <- MERINGUE::getSpatialNeighbors(rastPos, filterDist = 105)
plotNetwork(rastPos, w)
```

The adjacency weight matrix `w` contains the spatial relationship of the 384 hexagonal pixels. Thus, each point in the plot represents a hexagonal pixel. A line is drawn if the cells within that pixel are spatial neighbors with cells within another pixels as defined by the `filterDist` of 105. The `filterDist` is the maximum Euclidean distance for which two cells cannot be considered neighbors. This value was set based on the rasterized resolution of 100 um.

We can now calculate Moran's I with the gene expression matrix and identify significantly spatially auto-correlated genes.

```{r}
## rasterized genes-by-cells matrix, with cell pixels at 100 um resolution
mat <- assays(rastGexp)$pixel
I <- MERINGUE::getSpatialPatterns(mat, w)
```

Based on Moran's I calculations, 139 out of the 155 genes were identified as significantly spatially auto-correlated genes.

```{r}
results.filter <- MERINGUE::filterSpatialPatterns(mat = mat,
                                                  I = I,
                                                  w = w,
                                                  adjustPv = TRUE,
                                                  alpha = 0.05,
                                                  minPercentCells = 0.05,
                                                  verbose = TRUE, 
                                                  details = TRUE)
```

We can sort these spatially auto-correlated genes based their Moran's I statistic, as a higher I statistic generally correlates with greater significance. Let's identify the top three genes with the highest Moran's I statistics and visualize their expression patterns using the `plotRaster` function in `SEraster`.

```{r}
## sort spatially auto-correlated genes with Moran's I statistic
sortedResults <- results.filter[order(results.filter$observed, decreasing = TRUE), ]
## select 3 SVGs with the highest Moran's I
sgenes <- rownames(sortedResults)
for (sg in head(sgenes, 3)) {
  ## plot gene expression pattern
  plt <- SEraster::plotRaster(rastGexp, 
                              feature_name = sg,
                              name = sg)
  show(plt)
}
```

We also can compute the spatial cross correlation matrix to group significantly spatially variable genes into primary spatial gene expression patterns.

```{r}
## create a duplicate filtered results with details = FALSE this time
results.filter2 <- MERINGUE::filterSpatialPatterns(mat = mat,
                                                  I = I,
                                                  w = w,
                                                  adjustPv = TRUE,
                                                  alpha = 0.05,
                                                  minPercentCells = 0.05,
                                                  verbose = TRUE)

## compute spatial cross correlation matrix
scc <- MERINGUE::spatialCrossCorMatrix(mat = as.matrix(mat[results.filter2,]), 
                                       weight = w)
```

```{r}
# Identify primary patterns
par(mfrow=c(2,2), mar=rep(2,4))
ggroup <- MERINGUE::groupSigSpatialPatterns(pos = rastPos, 
                                            mat = as.matrix(mat[results.filter2,]), 
                                            scc = scc, 
                                            power = 1, 
                                            hclustMethod = 'ward.D', 
                                            deepSplit = 2,
                                            zlim=c(-1.5,1.5))
```

<details>

<summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>

